---
title: "README"
output: md_document  
rmarkdown::github_document:
pandoc_args: --webtex
header-includes:
   - \usepackage{amsmath, amssymb, amsthm}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Particle MCMC for agent-based SIS model

This repository contains code to implement the methodology described in the paper "Bayesian Inference on Agent-based Models of Disease Transmission", by Um and Adhikari (2022+)

This package uses the primary functions from [`agents`](https://github.com/nianqiaoju/agents).

# Installation

The packages can be installed with the `devtools` package:

```{r, eval = FALSE}
library(devtools) 
devtools::install_github(repo='Seungha-Um/agentSIS') 
```

The package with the vignettes can be installed with 

```{r, eval = FALSE}
devtools::install_github(repo='Seungha-Um/skewBART', build_vignettes = TRUE) 
```

and then accessed by running `browseVignettes("skewBART")` (to reproduce our results, one will need to increase the number of MCMC samples). 

