---
title: "PMCMC for agent-based SIS model"
author:  Seungha Um
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{PMCMC for agent-based SIS model}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
  editor_options: 
  chunk_output_type: console
---

# Introduction 

This vignette demonstrates how to implement PMCMC for agent-base SIS model using the `agentSIS` package. The small MCMC iterations and particles are used to reduce computation time; in practice, larger values (say, 20000 MCMC iterations and 300 particles) should be used. 

First, we load the requisite packages:

```{r, eval = TRUE, include=TRUE}
## Load library
library(tidyverse) # Load the tidyverse, mainly for ggplot
library(kableExtra) # For fancy tables
library(agentSIS) # Our package
library(zeallot) 
library(gmm)
options(knitr.table.format = 'markdown')
```

# Setup

The following code creates a function to simulate data from the model.

```{r, eval = TRUE, include=TRUE}
N <- 50;
steps <- 30;
X <- matrix(nrow = 2, ncol = N);
X[1,] <- 1;
X[2,] <- rnorm(N);

b_lam0 <- -1; b_lam1 <- 2; 
b_gam0 <- -1; b_gam1 <- -1;
Lambda <- fun_rate(b_lam0, b_lam1);
Gamma <- fun_rate(b_gam0, b_gam1);
Alpha <- rep(0.1, N)
Rho <- 0.8

## simulate observations
true_config <- list(N = N, alpha0 = Alpha, lambda = Lambda,
                    gamma = Gamma, rho = Rho)


sis_simulate <- function (days, true_config) {
  agent_states <- matrix(NA, nrow = true_config$N, ncol = 1 + days)
  agent_states[,1] <- (runif(N) < true_config$alpha0)
  for (d in c(2:(days + 1))) {
    alpha_t <- rep(NA, N)
    alpha_t <- sis_get_alpha(agent_states[, d - 1], true_config)
    agent_states[, d] <- (runif(N) < alpha_t)
  }
  y <- apply(agent_states, 2, function(xx) rbinom(n = 1, size = sum(xx), 
                                                  prob = true_config$rho))
  return(list(agent_states = agent_states, y = y))
}
```

We then create a dataset
```{r, eval = TRUE, include=TRUE}
set.seed(2022)
complete_data <- sis_simulate(days = steps, true_config = true_config);
y <- complete_data$y;
## reject the process if observations are zero
y
```

Next, we create an object containing the hyperparameters for the model using the `Hypers` function. This function uses default hyperparameters unless specific hyperparameters are provided by the user; see `?Hypers` for more details.

```{r, eval = TRUE, include=TRUE}
hypers <- Hypers()
```



The function `Opts` produces a list of the parameters for running the Markov
chain for agent-based SIS model. For illustration we use only 250 burn-in and save iterations; the defaults are 5000 for each. `Opts` can also be used to control some features of the chain, such as step_size for MHRW and the number of particles; see `?Opts` for more details.

```{r, eval = TRUE, include=TRUE}
opt <- Opt(num_burn = 250, num_save = 250)
```

# Implementing PMCMC for agent-based model 

After creating the hyperparameters and options, we fit the simulate data by running

```{r, eval = TRUE, include=TRUE}
fit <- PMCMC_SIS(X, y, opt=opt)
```








