---
title: "PMCMC for agent-based SIS model"
author:  Seungha Um
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{PMCMC for agent-based SIS model}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
  editor_options: 
  chunk_output_type: console
---

# Introduction 

This vignette demonstrates how to implement particle particle Markov-chain Monte Carlo (PMCMC) for disease transmission using the `agentSIS` package. The small MCMC iterations and particles are used to reduce computation time; in practice, larger values (say, 20000 MCMC iterations and 300 particles) should be used. 


# Statistcal Agent-based SIS Model

Agent-based model (ABM) is a simulation based modeling technique that aims to describe complex dynamic processes with a bottom-up approach. ABM provides considerable flexibility by explaining the complex dynamic process using simple rules that incorporate characteristics of individual entities (agents) and their interaction. For example, infectious disease is transmitted by the interaction between neighbors and by the attributes of each agent whose states fall into susceptible or infected states. 

In the disease transmission, the observed prevalence at time $t$ can be modeled as  
\begin{equation}\label{model:bin}
    y_t \vert I_t,\rho \sim \operatorname{Binomial}(I_t,\rho)
\end{equation}
where the detection probability $\rho$ accounts for the under-reported prevalence.

The observed prevalence at time $t$ depends on unobserved latent agent states $\mathbf{X}_t = (X_t^1, \ldots, X_t^N)$, which can be modeled as an independent Bernoulli distribution;
\begin{equation}\label{model:ber}
    X_t^n \ \vert \ \xi_{t-1}^n \sim \operatorname{Beroulli}(\xi_{t-1}^n).
\end{equation}
where $\xi_{t-1}^n$ corresponds to the transition probability between susceptible and infected state for agent $n$. Define agent-specific initial infection rate $\alpha_0^n$, infection rate $\lambda^n$ and recovery rate $\gamma^n$ as 
\begin{align*}
    \alpha_0^n = \left(1 + \exp(-\beta_{\alpha_0}^Tz^n)\right)^{-1}, \quad
    \lambda^n = \left(1 + \exp(-\beta_\lambda^Tz^n)\right)^{-1}, \quad
    \gamma^n = \left(1+\exp(-\beta_\gamma^Tz^n)\right)^{-1}
\end{align*}
where $\beta_{\alpha_0},\beta_\lambda,\beta_\gamma\in\mathbb{R}^d$ are parameters and $z^n\in\mathbb{R}^d$ are the covariates of agent $n$. 


Given the initial transition probability $\xi_0^n=\alpha_0^n$, the latent state $X_t$ for $t=2,\ldots,T$ evolves according to a Markov process with transition probability \abovedisplayskip14pt\belowdisplayskip14pt
\begin{align*}
    \xi^{n}_{t-1}= \begin{cases}\lambda^{n} \mathcal{D}(n)^{-1} \sum_{m \in \mathcal{N}_n} X_{t-1}^{m}, & \text { if } x_{t-1}^{n}=0 \\ 1-\gamma^{n}, & \text { if } x_{t-1}^{n}=1\end{cases}
\end{align*}
where $\mathcal{N}_n$ is a neighborhood for agent $n$ and $\mathcal{D}(n)$ is the number of neighbors of agent $n$. This transition probability depends only on the last value of the state of agents which is defined as a first-order Markovian. Since attributes of agents $z^n$ account for the infection and recovery rate, the transition probability $\xi_{t-1}^n$ is defined uniquely for each agent. 

While the underlying dynamics of the disease transmission are governed by agent states $\mathbf{X}_t$, these states are not observed directly. For example, public health agencies only report aggregate data at the population level rather than at the individual level to protect individuals' privacy; the number of infected agents is reported on a daily basis in COVID-19 data. As a result, the hidden Markov model represents our best understanding of ABM dynamics where the individual agent's states are treated as hidden and estimated using filtering approaches such as particle filter (PF). 



This figure highlights that it is suitable to treat agent states as hidden states and the heterogeneous transition rates between agents cannot be ignored whereas aggregate counts are observed.
 
 
 
 
 
## Load packages

First, we load the requisite packages:

```{r, eval = TRUE, include=TRUE}
## Load library
library(tidyverse) # Load the tidyverse, mainly for ggplot
library(kableExtra) # For fancy tables
library(agentSIS) # Our package
library(zeallot) 
library(reshape2)
options(knitr.table.format = 'markdown')
```

## Setup

The following code creates a function to simulate data from the model.

```{r, eval = TRUE, include=TRUE}
set.seed(2022)
N <- 30;
steps <- 30;
X <- matrix(nrow = 2, ncol = N);
X[1,] <- 1;
X[2,] <- rnorm(N);

b_lam0 <- -1; b_lam1 <- 2; 
b_gam0 <- -1; b_gam1 <- -1;
Lambda <- fun_rate(b_lam0, b_lam1);
Gamma <- fun_rate(b_gam0, b_gam1);
Alpha <- rep(0.1, N)
Rho <- 0.8

## simulate observations
true_config <- list(N = N, alpha0 = Alpha, lambda = Lambda,
                    gamma = Gamma, rho = Rho)


sis_simulate <- function (days, true_config) {
  agent_states <- matrix(NA, nrow = true_config$N, ncol = 1 + days)
  agent_states[,1] <- (runif(N) < true_config$alpha0)
  for (d in c(2:(days + 1))) {
    alpha_t <- rep(NA, N)
    alpha_t <- sis_get_alpha(agent_states[, d - 1], true_config)
    agent_states[, d] <- (runif(N) < alpha_t)
  }
  y <- apply(agent_states, 2, function(xx) rbinom(n = 1, size = sum(xx), 
                                                  prob = true_config$rho))
  return(list(agent_states = agent_states, y = y))
}
```

We then create a dataset
```{r, eval = TRUE, include=TRUE}
complete_data <- sis_simulate(days = steps, true_config = true_config);
y <- complete_data$y;
```

Next, we create an object containing the hyperparameters for the model using the `Hypers` function. This function uses default hyperparameters unless specific hyperparameters are provided by the user; see `?Hypers` for more details.

```{r, eval = TRUE, include=TRUE}
hypers <- Hypers()
```

The function `Opts` produces a list of the parameters for running the Markov
chain for agent-based SIS model. For illustration we use only 250 burn-in and save iterations; the defaults are 5000 for each. `Opts` can also be used to control some features of the chain, such as step_size for MHRW and the number of particles; see `?Opts` for more details.

```{r, eval = TRUE, include=TRUE}
opt <- Opt(num_burn = 100, num_save = 100)
```

## Genterated data

```{r, eval = TRUE, fig.width=5, fig.height=4.5, fig.align='center'}
## genrate figure relecting hidden states and observations
A <- complete_data$agent_states
longData <- melt(A)
ind <- longData$value == TRUE

longData$status <- "Susceptible"
longData$status[ind] <- "Infected"
longData$status <- factor(longData$status, levels=c("Susceptible", "Infected"))

df <- data.frame(x=c(1:length(y)), y = y)

ggplot(longData, aes(x = Var2, y=Var1)) + 
  geom_tile(aes(fill = status, width=0.8, height=0.8), size=2) + 
  scale_fill_manual(values=c("grey90","grey60")) + 
  geom_line(data=df, aes(x = x, y = y), size=1) + 
  xlab("time") + ylab("agents") + 
  theme_bw() + 
  theme(legend.position = "bottom", legend.title=element_blank(), 
        panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank()) 

```

The figure represents simulated hidden states and observations from agent-based SIS model. Black line represents observed infection counts and each grid represents agent state which is hidden. The color scale of grid indicates agent state (susceptible or infected) and the agent indices (y-axis) are sorted by magnitude of infection rates. This figure highlights that it is suitable to treat agent states as hidden states and the heterogeneous transition rates between agents cannot be ignored whereas aggregate counts are observed.

## Implementing PMCMC for agent-based model 

For PMCMC, there are PHHM (MH scheme) and Particle Gibbs (Gibbs sampling scheme). To improve the performance of Particle Gibbs (PG), we can add the ancestor sampling step. Our package can implement PMMH, PG and PG with ancestor sampling. 

The functions create hyperparameters and options by defaults if not provided. We fit the simulate data by running

```{r, eval = TRUE, include=TRUE}
fit_PMMH <- PMMH_SIS(X, y, opt = opt)
fit_PG <- PG_SIS(X, y, opt = opt)
fit_PG_AS <- PG_AS_SIS(X, y, opt = opt)
```

We can compare three methods in terms of the estimated infection counts over time.

```{r, eval = TRUE, include=TRUE}
est_mat <- data.frame(time = rep(c(0:steps), 4), 
                      est = c(y, fit_PMMH$obs_counts_mean, fit_PG$obs_counts_mean,
                              fit_PG_AS$obs_counts_mean), 
                      grp = rep(c("observation","PMMH", "PG", "PGAS"), 
                                each=length(fit_PG$obs_counts_mean)))
```

```{r, eval = TRUE, fig.width=6, fig.height=4, fig.align='center'}
ggplot(est_mat) + 
  geom_line(aes(x = time, y = est, color = grp, linetype = grp)) + theme_bw() +
  scale_color_manual(values=c("#999999", "#E69F00", "#117733", "#CC6677")) + 
  scale_linetype_manual(values=c("dashed","solid","solid","solid")) + 
  theme(legend.title=element_blank(),legend.position="bottom") + ylab("estimated counts")

```

Posterior mean with 95\% CI from PMMH

```{r, eval = TRUE, include=TRUE}
table_PMMH <- rbind(post_mean = colMeans(fit_PMMH$param_mat),
apply(fit_PMMH$param_mat, 2, function(x) quantile(x, c(0.025, 0.975))))
colnames(table_PMMH) <-  c("$\\alpha_0$", "$\\alpha_1$", "$\\lambda_0$", "$\\lambda_1$","$\\gamma_0$","$\\gamma_1$","$\\rho$")
kable(table_PMMH)
```

Posterior mean with 95\% CI from PG
```{r, eval = TRUE, include=TRUE}
table_PG <- rbind(post_mean = colMeans(fit_PG$param_mat),
apply(fit_PG$param_mat, 2, function(x) quantile(x, c(0.025, 0.975))))
colnames(table_PG) <-  c("$\\alpha_0$", "$\\alpha_1$", "$\\lambda_0$", "$\\lambda_1$","$\\gamma_0$","$\\gamma_1$","$\\rho$")
kable(table_PG)
```

Posterior mean with 95\% CI from PG with ancestor sampling
```{r, eval = TRUE, include=TRUE}
table_PG_AS <- rbind(post_mean = colMeans(fit_PG_AS$param_mat),
apply(fit_PG_AS$param_mat, 2, function(x) quantile(x, c(0.025, 0.975))))
colnames(table_PG_AS) <- c("$\\alpha_0$", "$\\alpha_1$", "$\\lambda_0$", "$\\lambda_1$","$\\gamma_0$","$\\gamma_1$","$\\rho$")
kable(table_PG_AS)
```





